Wrattler: \textnormal{A platform for AI-assisted data science}

### 2018-02-28 report/paper.tex
Minor tweaks throughout to make layout nicer

\section{Introduction}
Data science is an iterative, exploratory process that requires a collaboration between a
computer system and a human. A computer can provide advice based on statistical analysis of the
data and discover hidden structures or corner cases, but only a human can decide what those mean
and decide how to handle them in data science scripts. Data science is often cited as an expensive
and time consuming task, especially due the costs of data cleaning and data wrangling.
We propose four fundamental reasons why
practical data science is expensive:

@@

\vspace{1em}
To meet these challenges, we present Wrattler,
a new type of system for data science
that aims to transform the process
of data analysis.
Wrattler combines the interactive and literal programming paradigms
of notebook systems such as Jupyter
with new advances in AI systems for data wrangling and in provenance.
The main design principles in Wrattler are:

\paragraph{Interactive.} Interactivity
is a necessary because ``big data is big'', so the analyst learns about
the data set and the problem as she explores it.
Wrattler enables an efficient interaction by bringing computation closer to the human.
Notebooks run in the browser, cache partial results of computations and provide previews
of script results on-the-fly during development.

@@

\begin{figure}
\vspace{-0.5em}
\includegraphics[scale=0.6]{diagram.pdf}
\vspace{-0.5em}
\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
  Wrattler, those functions are separated and enriched with AI assistants.}}
\label{fig:arch}
\vspace{-0.5em}
\end{figure}

The architecture of Wrattler allows us to address these issues, as well as to provide a platform
for building novel AI assistants and interactive programming. The architecture is illustrated
in Figure~\ref{fig:arch}. The components of Wrattler are:
%
\paragraph{Data store.} Imported external data, results of running scripts and of
applying AI assistants are stored in the data store. It stores versioned data frames with
metadata such as types, inferred semantics, data formats or provenance.

\paragraph{Language runtimes.} Scripts are evaluated by one or more language runtimes.
The runtimes read input data from and write results back to the data store.

\paragraph{AI assistants.} When invoked from the notebook, AI assistants read data
from data store and provide hints to the analyst. They help to write data cleaning
scripts or annotate data with additional metadata such as inferred types.

\paragraph{Notebook.} The notebook is displayed in a web browser and orchestrates
all other components. The browser builds a dependency graph between cells or individual
expressions in the cells. It calls language runtimes to evaluate code that has changed,
AI assistants to provide hints and reads data from the data store to display results.

@@

Wrattler consists of a notebook user interface running in the web browser, which communicates with
a number of server-side components including language runtimes, the data store and AI assistants.
In this section, we discuss the components in more detail, starting with the dependency graph
that is maintained on the client-side, by the web browser. AI assistants are discussed later
in Section~\ref{sec:ai}

@@

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
\textit{\small{We retrieve data from a government web site using Python:}}\\[0.3em]
\quad\ident{url}=\ident{open}(\str{http://data.gov.uk/\ldots/bb2014.csv})\\
\quad\ident{bbRaw}=\ident{csv.reader}(\ident{url})\\[0.2em]
\textit{\small{In the next block, we used an AI assistant to assist with cleaning}}\\[-0.1em]
\textit{\small{and created a script that marks \#!NULL values as missing:}}\\[0.3em]
\quad\kvd{let}~\ident{bbClean} = \ident{bbRaw}.\ident{missing}.\ident{as\_missing}(\str{\#!NULL})\\[0.3em]
\textit{\small{Finally, we use JavaScript to create a histogram of upload speeds:}}\\[0.3em]
\quad\ident{vega}.\ident{hsit}(\ident{bbClean}, \str{Upload speed})
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{\small{A small notebook consisting of three cells. The first Python cell exports
\ident{bbRaw}; the second Wrattler cell uses it and exports \ident{bbClean} and the
JavaScript cell visualizes \ident{bbClean}.}}
\label{fig:notebook}
\end{figure}
\begin{figure}
\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}

@@

Language runtimes are responsible for evaluating code and assisting with code analysis during
the construction of dependency graph. They can run as server-side components (e.g.~for R and Python)
or as client-side components (for JavaScript and small analytical DSLs used by AI assistants).
Unlike Jupyter kernels, a language runtime is stateless. It reads data from and writes data to
the data store. (Using a cache for efficiency.)

@@

More specifically, an AI assistant defines a domain specific language (DSL) for the task at hand. When invoked from
a notebook, it guides the user in creating a script in the DSL. Running the script performs
the required operation (such as data cleaning). The assistant defines the semantics
of those operations and, when the script is executed, stores the result into a data
store. This new data frame can contain new data or additional information (such as their
inferred probabilistic type).
The interaction is illustrated in Figure~\ref{fig:arch}. An AI assistant reads data from the data
store and returns suggestions to the notebook, which the user can review and accept.

@@

@@

In the above example, the interaction is provided through the ``.'' operator of the scripting
language. When the analyst types the first line to invoke the AI assistant and types ``.''
the datadiff assistant offers the most likely patches and the user can choose which ones
to accept. Scripts constructed in such way are readable, allowing the data analyst to look at them
and verify that the transformations accomplish the intended task (such verification further
benefits from live previews that Wrattler provides when the analyst navigates through the
individual transformations).

@@

\vspace{1em}
The Wrattler architecture supports the development of AI assistants in a number of ways. First, AI
assistants have access to the data store and can use it to provide guidance about both data and
code. Second, Wrattler includes an extensible framework for defining domain-specific languages
(DSLs) that can describe, for example, data extraction, cleaning, transformation or visualization.
The data store provides AI assistants with
two sources of information that they can exploit
to determine what guidance to provide the analyst:
%
\begin{itemize}
\item[--] AI assistants can access all data stored in the data
store. This includes imported external data files, such as raw text or raw HTML content from which
the analyst wants to extract data, as well as structured data frames imported from CSV files or
produced as a result of previous computations. Moreover, an AI assistant can also access multiple
versions of a file.

@@

Assuming the data frame is extracted from a government web page, each row of the data set
could further be annotated with the URL of the data source. This makes it possible to track
provenance at a fine-grained level. If we then append two such data frames, we will know
which row comes from which government data set.

>>>>

\section{Introduction}
Data science is an iterative, exploratory process that requires a collaboration between a
computer system and a human. A computer can provide advice based on statistical analysis of the
data and discover hidden structures or corner cases, but only a human can decide what those mean
and decide how to handle them in scripts. Data science is often cited as an expensive
and time consuming task, especially due the costs of data cleaning and data wrangling.
We propose four fundamental reasons why
practical data science is expensive:

@@

\vspace{1em}
\noindent
To meet these challenges, we present Wrattler,
a new type of system for data science
that aims to transform the process
of data analysis.
Wrattler combines the interactive and literal programming paradigms
of notebook systems such as Jupyter
with new advances in AI systems for data wrangling and in provenance.
The main design principles are:

\paragraph{Interactive.} Interactivity
is a necessary because ``big data is big''. The analyst learns about
the data set and the problem as she explores it.
Wrattler enables an efficient interaction by bringing computation closer to the human.
Notebooks run in the browser, cache partial results of computations and provide previews
of script results on-the-fly.

@@

The architecture of Wrattler allows us to address these issues, as well as to provide a platform
for building novel AI assistants and interactive programming. The architecture is illustrated
in Figure~\ref{fig:arch}. The components of Wrattler are:

\vspace{-0.25em}
\paragraph{Data store.} Imported external data, results of running scripts and of
applying AI assistants are stored in the data store. It stores versioned data frames with
metadata such as types, inferred semantics, data formats or provenance.

\vspace{-0.25em}
\paragraph{Language runtimes.} Scripts are evaluated by one or more language runtimes.
The runtimes read input data from and write results back to the data store.

\vspace{-0.25em}
\paragraph{AI assistants.} When invoked from the notebook, AI assistants read data
from data store and provide hints to the analyst. They help to write data cleaning
scripts or annotate data with additional metadata such as inferred types.

\vspace{-0.25em}
\paragraph{Notebook.} The notebook is displayed in a web browser and orchestrates
all other components. The browser builds a dependency graph between cells or individual
expressions in the cells. It calls language runtimes to evaluate code that has changed,
AI assistants to provide hints and reads data from the data store to display results.

@@

Wrattler consists of a user interface running in the web browser, which communicates with
a number of server-side components. In this section, we discuss the components in more 
detail, starting with the dependency graph that is maintained on the client-side, by the 
web browser. AI assistants are discussed later in Section~\ref{sec:ai}.

@@

\begin{figure}
\includegraphics[scale=0.6]{diagram.pdf}
\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
  Wrattler, those functions are separated and enriched with AI assistants.}}
\label{fig:arch}
\end{figure}

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
\textit{\small{In the first block, we retrieve data from a government web site.}}\\[-0.1em]
\textit{\small{We use Python to download the data and parse it as a CSV file.}}\\[0.3em]
\quad\ident{url}=\ident{open}(\str{http://data.gov.uk/\ldots/bb2014.csv})\\
\quad\ident{bbRaw}=\ident{csv.reader}(\ident{url})\\[0.2em]
\textit{\small{In the next block, we used an AI assistant to assist with cleaning}}\\[-0.1em]
\textit{\small{and created a script that marks \#!NULL values as missing:}}\\[0.3em]
\quad\kvd{let}~\ident{bbClean} = \ident{bbRaw}.\ident{missing}.\ident{as\_missing}(\str{\#!NULL})\\[0.3em]
\textit{\small{Finally, we use JavaScript to create a histogram of upload speeds:}}\\[0.3em]
\quad\ident{vega}.\ident{hsit}(\ident{bbClean}, \str{Upload speed})
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{\small{A small notebook consisting of three cells. The first Python cell exports
\ident{bbRaw}; the second Wrattler cell uses it and exports \ident{bbClean} and the
JavaScript cell visualizes \ident{bbClean}.}}
\label{fig:notebook}
\end{figure}

@@

Language runtimes are responsible for evaluating code and assisting with code analysis during
the construction of dependency graph. They can run as server-side components (e.g.~for R and Python)
or as client-side components (in case of JavaScript).
Unlike Jupyter kernels, a language runtime is stateless. It reads data from and writes data to
the data store. (Using a cache for efficiency.)

@@

More specifically, an AI assistant defines a domain specific language (DSL) for the task at hand. When invoked from
a notebook, it guides the user in creating a script in the DSL. Running the script performs
the required operation (such as data cleaning). The assistant defines the semantics
of those operations and, when the script is executed, stores the result into a data
store. This new data frame can contain new data or additional information (such as their
inferred types).
The interaction is illustrated in Figure~\ref{fig:arch}. An AI assistant reads data from the data
store and returns suggestions to the notebook, which the user can review and accept.

@@

\begin{figure}
\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}

@@

In the above example, the interaction is provided through the ``.'' operator of the scripting
language. When the analyst types the first line to invoke the AI assistant and types ``.''
the datadiff assistant offers the most likely patches and the user can choose which ones
to accept.

Scripts consisting of a chain of simple method calls are readable, allowing the data analyst to look at them
and verify that the transformations accomplish the intended task (such verification further
benefits from live previews that Wrattler provides when the analyst navigates through the
individual transformations).

@@

\vspace{1em}
\noindent
The Wrattler architecture supports the development of AI assistants in a number of ways. First, AI
assistants have access to the data store and can use it to provide guidance about both data and
code. Second, Wrattler includes an extensible framework for defining domain-specific languages
(DSLs) that can describe, for example, data extraction, cleaning, transformation or visualization.

The data store provides AI assistants with
two sources of information that they can exploit
to determine what guidance to provide the analyst:
%
\begin{itemize}
\item[--] AI assistants can access all data stored in the data
store. This includes imported external data files, such as raw text or raw HTML content from which
the analyst wants to extract data, as well as structured data frames imported from CSV files or
produced as a result of previous computations. Moreover, an AI assistant can also access multiple
versions of a file.

@@

\noindent
Assuming the data frame is extracted from a government web page, each row of the data set
could further be annotated with the URL of the data source. This makes it possible to track
provenance at a fine-grained level. If we then append two such data frames, we will know
which row comes from which government data set.
